Probing heavy ion collisions using quark and gluon jet substructure

### 2020-08-27 qg_ML.tex
updated draft with separate appendix for model details and added figures for networks

We combine the five jet substructure observables using a multivariate model implemented in Keras \cite{keras}, with two hidden dense layers of 10 nodes each. We use the rectified linear unit (ReLU) \cite{nair2010rectified} and sigmoid activation functions for the dense layers and the output layer, respectively. The model training utilizes the Adam optimizer \cite{adam} and the binary cross entropy loss function. We perform cross validation using Monte Carlo samples which are split into two random halves for training and testing.

@@

A deep convolutional neural network is implemented in Keras with the TensorFlow \cite{DBLP:journals/corr/AbadiBCCDDDGIIK16} backend. The neural network model consists of three convolution layers of sequentially reducing sizes, each with 8 filters followed by a max-pooling layer. The convolution layers use a hyperbolic tangent activation function, and the output layer uses a sigmoid activation function. The output layer is preceded by three layers, a dropout layer with a rejection score of 0.20, a dense layer with 20 nodes and an additional dropout layer with a reduced rejection score of 0.10. The additional dropout layer helps filtering less important features from previous layers and thus increases the speed and efficiency of the model training. The model is trained using the Adam optimizer with binary cross entropy loss function.

@@

\newpage
\appendix
\section*{Appendix}
\section{Probing subjet energy flows with telescoping deconstruction}

@@

A neural network consisting of three dense layers with 100 nodes each is trained on the telescoping deconstruction observables up to T$N$ order. The training can be performed on a typical laptop CPU in fewer than five minutes. One could also use boosted decision trees (BDT) to combine variables \cite{Chien:2017xrb}. We compare the performance of our method to jet images using a CNN as done in the main text. The jet images are size $33\times 33$ and span a $2R\times 2R$ patch of the rapidity-azimuth plane with the intensity of each pixel corresponding to the total $p_T$ of particles in the pixel. The jet images are pre-processed and standardized according to the procedure in \Ref{Komiske:2016rsd}. The CNN architecture consists of three 48-filter convolutional layers with filter sizes of $8\times 8$, $4\times 4$, and $4\times 4$ followed by a 128-unit dense layer and a 2-unit softmaxed output layer. A $2\times 2$ maxpooling is performed after each convolutional layer with a stride length of 2. The dropout rate was taken to be 0.1 for all layers. All neural networks are implemented using the Python deep learning library Keras~\cite{keras} with the Theano backend~\cite{bergstra2010theano}. Rectified linear unit (ReLU) activation functions~\cite{nair2010rectified} and He-uniform model weight initialization~\cite{heuniform} are used. The networks are trained using the Adam algorithm~\cite{adam} with a learning rate of $10^{-3}$ and a batch size of 256 for 50 epochs with a patience parameter of 8, and the best model is selected based on validation set performance.

>>>>

We combine the five jet substructure observables using a multivariate model implemented in Keras \cite{keras}. The details of the model description are given in Appendix~\ref{app:models}.

@@

A deep convolutional neural network is implemented in Keras~\cite{DBLP:journals/corr/AbadiBCCDDDGIIK16}. The neural network architecture and the corresponding activation functions are detailed in the appendix.

@@

\newpage
\appendix
\section*{Appendix}

\section{Description of the MLP and CNN network architectures}
\label{app:models}
The MLP utilized in this classification analysis utilizes two hidden dense layers of 10 nodes each as shown in Fig~\ref{fig:mlp}. We use the rectified linear unit (ReLU) \cite{nair2010rectified} and sigmoid activation functions for the dense layers and the output layer, respectively. The model training utilizes the Adam optimizer \cite{adam} and the binary cross entropy loss function. We perform cross validation using Monte Carlo samples which are split into two random halves for training and testing.

\begin{figure}[t]
\centering
\includegraphics[width=0.6\columnwidth]{FigMLP.png}
\caption{Schematic of the MLP utilized for physics observables where the two hidden layers are shown and particular nodes that are not linked are dropped (shown as an example) during training.}
\label{fig:mlp}
\end{figure}

The CNN architecture consists of three 48-filter convolutional layers with filter sizes of $8\times 8$, $4\times 4$, and $4\times 4$ followed by a 128-unit dense layer and a 2-unit softmaxed output layer. The network architecture is schematically shown in Fig~\ref{fig:dcnn}. A $2\times 2$ maxpooling is performed after each convolutional layer with a stride length of 2. The dropout rate was taken to be 0.1 for all layers. All neural networks are implemented using the Python deep learning library Keras~\cite{keras} with the Theano backend~\cite{bergstra2010theano}. Rectified linear unit (ReLU) activation functions~\cite{nair2010rectified} and He-uniform model weight initialization~\cite{heuniform} are used. The networks are trained using the Adam algorithm~\cite{adam} with a learning rate of $10^{-3}$ and a batch size of 256 for 50 epochs with a patience parameter of 8, and the best model is selected based on validation set performance.

\begin{figure}[t]
\centering
\includegraphics[width=.8\columnwidth]{FigDCNN_wolabels.png}
\caption{Schematic of the DCNN with maxpooling layers (not shown) located in-between each convolution layer. The output layer for classification is taken as a result of a dense layer (not shown).}
\label{fig:dcnn}
\end{figure}

\section{Probing subjet energy flows with telescoping deconstruction}

@@

A neural network consisting of three dense layers with 100 nodes each is trained on the telescoping deconstruction observables up to T$N$ order. The training can be performed on a typical laptop CPU in fewer than five minutes. One could also use boosted decision trees (BDT) to combine variables \cite{Chien:2017xrb}. We compare the performance of our method to jet images using a CNN as done in the main text. The jet images are size $33\times 33$ and span a $2R\times 2R$ patch of the rapidity-azimuth plane with the intensity of each pixel corresponding to the total $p_T$ of particles in the pixel. The jet images are pre-processed and standardized according to the procedure in \Ref{Komiske:2016rsd}.
